<!--
%\VignetteEngine{knitr::knitr}
%\VignetteIndexEntry{Doing more with dendrogram objects using dendextend}
-->

Doing more with dendrogram objects
==================================
A tutorial on using the **dendextend** package
================================================


Introduction
-------------


### The "dendrogram" object

The "dendrogram" class provides general functions for handling tree-like structures in R. It is intended as a replacement for similar functions in hierarchical clustering and classification/regression trees, such that all of these can use the same engine for plotting or cutting trees.

Dendrogram are basically list


However, many basic features are still missing from the dendrogram class.

### Motivation for creating {dendextend}
The **{dendextend}** package aims at filling some gaps in base R, by extending the available functions for dendrogram manipulation, statistical analysis, and visualization.

This vignettes Provides a step-by-step description of the functionality provided by the **dendextend** package.


### Installing {dendextend}

To install the stable version from CRAN use:

```r
# install.packages('dendextend') # not yet available from CRAN
```

To install the GitHub version use:

```r
if (!require('devtools')) install.packages('devtools'); require('devtools')
install_github('dendextend', 'talgalili')
```











This is an example of Markdown vignettes in R. Before R 3.0.0, only Sweave/PDF vignettes were supported in R. Markdown is gaining popularity over the years due to its simplicity, and R 3.0.0 starts to support package vignettes written in [R Markdown](http://www.rstudio.com/ide/docs/authoring/using_markdown).

## Package vignettes

To enable Markdown vignettes in an R package, you need to

- add `*.Rmd` files under the `vignettes` directory
- add `VignetteBuilder: knitr` to the `DESCRIPTION` file
- specify the vignette engine `\VignetteEngine{knitr::knitr}` in the `Rmd` files (inside HTML comments)

## View vignettes

And R will load the **knitr** package to build these vignettes to HTML files, and you can see them when you open the HTML help:

```{r eval=FALSE}
help(package = 'YourPackage', help_type = 'html')
```

## Examples

Below are some code chunks as examples.

```{r hello, results='asis'}
cat('_hello_ **markdown**!', '\n')
```

Normally you do not need any chunk options.

```{r}
1+1
10:1
rnorm(5)^2
strsplit('hello, markdown vignettes', '')
```

Feel free to draw beautiful plots and write math $P(X>x)=\alpha/2$.

```{r}
n=300; set.seed(123)
par(mar=c(4,4,.1,.1))
plot(rnorm(n), rnorm(n), pch=21, cex=5*runif(n), col='white', bg='gray')
```

You can use your own CSS file instead of the built-in style in the **markdown** package -- just set the option `markdown.HTML.stylesheet`, e.g.

```{r css, eval=FALSE}
options(markdown.HTML.stylesheet = 'path/to/a/custom/style.css')
```

For this vignette, I use the `markdown.HTML.header` option to modify a few CSS rules in the default CSS style:

```{r header}
options(markdown.HTML.header = system.file('misc', 'vignette.css', package='knitr'))
```
